TSELM: Target speaker extraction using discrete tokens and language models

### 2024-09-09 src/paper.tex
update

In contrast to blind speech separation, which seeks to isolate individual utterances from a mixture of known speakers, target speaker extraction (TSE) focuses on extracting only the voice of the target speaker using auxiliary information. Current models are predominantly discriminative, employing masking strategies to directly minimize the distance between the estimated and clean speech signals \cite{luo2019conv,spex_plus,sepformer,sef_net}. However, these discriminative approaches often struggle to generalize to unseen data and may introduce undesirable distortions \cite{distortion}. To address these limitations, generative models have been proposed. These models aim to learn the underlying distribution of the target speech, rather than mapping mixed speech to clean speech. Generative models have been shown to achieve performance comparable to that of discriminative models \cite{target_diff,tokensplit} in blind speech separation and TSE.

The discretization of audio has gained significant attention with the advancement of language models (LMs). This method converts audio into discrete tokens, allowing LMs to model them, thereby simplifying audio generation tasks by transforming complex regression problems into classification tasks \cite{dasb}. SSL models such as HuBERT \cite{hubert} and WavLM \cite{wavlm} have demonstrated outstanding performance across numerous downstream tasks \cite{superb}, as they extract continuous representations rich in semantic and timbral information from speech. As shown in \cite{dasb}, SSL models outperform neural audio codecs like \cite{dac} in tasks like speech enhancement and separation. Consequently, this paper primarily focuses on exploring the discretization of SSL models.

@@

In the encoding stage, both reference and mixed speech are tokenized using WavLM and Kmeans as the tokenizer. Unlike SELM \cite{selm}, which utilizes the 6th layer of WavLM as input to the Kmeans layer, we follow the approach in \cite{dasb} and use outputs from WavLM's hidden layers 1, 3, 7, 12, 18, and 23 as input to 6 separate Kmeans models, one for each layer. Our results show that using multiple layers as input is superior to using only the 6th layer. The reference speech is passed directly to the encoder, while for mixed speech, we concatenate the reference speech to both sides of the mixture before passing it through the WavLM model. After tokenization, we retain only the tokens corresponding to the mixed speech.
In the modeling stage, an attention embedding mechanism is employed to combine the embeddings from all layers. A cross-attention mechanism, similar to that used in \cite{usef_tes}, is applied to inject speaker-specific information. An encoder-only language model, followed by a linear classifier, is then used to generate the reconstructed tokens.
In the decoding stage, we leverage the pretrained scalable HiFi-GAN from \cite{unit_hifi} to reconstruct the audio from the discrete tokens. Unlike SELM \cite{selm}, where a conformer detokenizer is trained to reconstruct WavLM embeddings before passing them through HiFi-GAN, the scalable HiFi-GAN in \cite{unit_hifi} uses a dropout mechanism to directly reconstruct audio from multiple layers of tokens, eliminating the need for a conformer detokenizer and the complexity of training a separate HiFi-GAN for each layer. Both the encoder and decoder are kept frozen during training, and an overview of the model is shown in Fig.\ref{model}.
Extensive experiments demonstrate that our method achieves excellent speech quality and comparable intelligibility results. To the best of our knowledge, this is the first study to explore target speaker extraction using discrete tokens. Our demos are available at \href{https://beilong-tang.github.io/TSELM.demo/}{https://beilong-tang.github.io/TSELM.demo/}.

>>>>

In contrast to blind speech separation, which seeks to isolate individual utterances from a mixture of known speakers, target speaker extraction (TSE) focuses on extracting only the voice of the target speaker using auxiliary information. Current models are predominantly discriminative, employing masking strategies to directly minimize the distance between the estimated and clean speech signals \cite{luo2019conv,spex_plus,sepformer,sef_net}. However, these discriminative approaches often struggle to generalize to unseen data and may introduce undesirable distortions \cite{distortion}. To address these limitations, generative models aim to learn the underlying distribution of the target speech, rather than mapping mixed speech to clean speech. Recently, generative models have been shown to achieve performance comparable to that of discriminative models in blind speech separation \cite{tokensplit} and TSE \cite{target_diff}.

The discretization of audio has gained significant attention with the advancement of language models (LMs). This method converts audio into discrete tokens, and leverages LMs to model them, thereby simplifying audio generation tasks by transforming complex regression problems into classification tasks \cite{dasb}. SSL models such as HuBERT \cite{hubert} and WavLM \cite{wavlm} have demonstrated outstanding performance across numerous downstream tasks \cite{superb}, as they extract continuous representations rich in semantic and timbral information from speech. As shown in \cite{dasb}, SSL models outperform neural audio codecs \cite{dac} in tasks like speech enhancement and separation. Consequently, this paper primarily focuses on exploring the discretization of SSL models.

@@

In the encoding stage, both reference and mixed speech are tokenized using WavLM and Kmeans. The reference speech is passed directly to the encoder, while for mixed speech, we concatenate the reference speech to both sides of the mixture before passing it through the WavLM model. After tokenization, we retain only the tokens corresponding to the mixed speech.
In the modeling stage, an attention embedding mechanism is employed to combine the embeddings from all layers. A cross-attention mechanism, similar to that used in \cite{usef_tes}, is applied to inject speaker-specific information. An encoder-only language model, followed by a linear classifier, is then used to generate the reconstructed tokens.
In the decoding stage, we leverage the pretrained scalable HiFi-GAN from \cite{unit_hifi} to reconstruct the audio from the discrete tokens. Unlike SELM \cite{selm}, where a conformer detokenizer is trained to reconstruct WavLM embeddings before passing them through HiFi-GAN, the scalable HiFi-GAN in \cite{unit_hifi} uses a dropout mechanism to directly reconstruct audio from multiple layers of tokens, eliminating the need for a conformer detokenizer and the complexity of training a separate HiFi-GAN for each layer. Both the encoder and decoder are kept frozen during training, and an overview of the model is shown in Fig.\ref{model}.
Extensive experiments demonstrate that our method achieves excellent speech quality and comparable intelligibility results. To the best of our knowledge, this is the first study to explore target speaker extraction using discrete tokens. Our demos are available at \href{https://beilong-tang.github.io/TSELM.demo/}{https://beilong-tang.github.io/TSELM.demo/}.
